---
title: "Impact of Severe Weather Events Across the United States between 1950 and 2011"
author: "Lawrence C. Chen"
date: "Friday, May 23, 2015"
output:
  html_document:
    keep_md: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
---

# Synopsis

Although destructive weather events like [Hurricane Katrina](http://en.wikipedia.org/wiki/Hurricane_Katrina)
and the [Great Flood of 1993](http://en.wikipedia.org/wiki/Great_Flood_of_1993)
are infrequent, their resulting impact to the population health and damage to
the economy lasted months to years. A significant root causes that made the
impact of these events far worse can be attributed to the lack of preparation
and anticipation.

Analyzing records between 1950 and 2011 recorded in the [NOAA Storm Database](http://www.ncdc.noaa.gov/stormevents/ftp.jsp),
states in the Southeast and Southwest regions are most susceptible to weather
events commonly associated with wind and water elements such as hurricane, storm
and and flood, next by weather event associated with ice elements such as heavy
snow.  However, it is extreme weather events associated to high temperature and
drought (element heat) that most commonly lead to significant population/crop
damage and are most harmful to population health.


# Loading and preprocessing the raw data

```{r download and sample data, results='hide'}
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
filename <- 'repdata-data-StormData.csv.bz2'

#check if the data file is available
if (!file.exists(filename)) {
    download.file(url = url, destfile = filename)
}

sample <- read.csv(filename, nrows = 100)

colClasses <- vapply(sample, class, 'character')
ch <- grep('(logical|integer|factor)', colClasses)
colClasses[ch] <- 'character'
```


```{r data, cache=TRUE, message=FALSE}
library(data.table)

dts <- as.data.table(read.csv(filename, na.strings = '', colClasses = colClasses))
```

This data set consists of `r dim(dts)[1]` observations and `r dim(dts)[2]`
variables.  The first 6 rows of the data are as below:

```{r, echo=FALSE, message=FALSE}
library(data.table)

head(dts)
```

Columns crucial to the analysis are:

* STATE (2-letter state designation, including the US Territories)
* BGN_DATE (date which the event occurred)
* EVTYPE (type of event)
* FATALITIES (no. of human fatalities)
* INJURIES (no. of human injuries)
* PROPDMG (dollar amount of property damage)
* PROPDMGEXP (magnitude of amount in thousand (K), million (M), or billion (B))
* CROPDMG (dollar amount of crop damage)
* CROPDMGEXP (magnitude of amount in thousand (K), million (M), or billion (B))

## Data cleansing
```{r colnames, results='hide'}
library(data.table)

setnames(dts, c(1, 35), c('STATE_FIPS', 'LONGITUDE_E'))

dts[, ':='(STATE_FIPS =as.integer(STATE_FIPS),
           BGN_DATE   =as.IDate(BGN_DATE, '%m/%d/%Y'),
           EVTYPE     =tolower(EVTYPE),           
           REFNUM     =as.integer(REFNUM))]

obs.total <- dim(dts)[1]
obs.dmg <- dim(dts[PROPDMG > 0 | CROPDMG > 0])[1]
obs.badexp <- dts[PROPDMG > 0 | CROPDMG > 0,
                  sum((!grepl('[KMBkmb]', PROPDMGEXP) |
                           !grepl('[KBMkbm]', CROPDMGEXP)))]

```

Of `r obs.total` total observations (obs), `r obs.dmg` obs. contain property or
crop damage estimates.  However, `r round(obs.badexp / obs.dmg * 100, 2)` percent
of which do not have the proper magnitude designations: K, M or B. In those
instances, thousand (K) is assumed. If magnitude is specified as one of the number
between 1 to 9, the number is used as the exponent of 10.

### Adjusting property and crop damages

```{r damage calc, results='hide'}
dmgAmt <- function(value, magnitude = 'K') {
    #returns dollar value in mulitple of thousands.
    #If magnitude is not one of 'K', 'M' or 'B', assume 'K'.

    #assumes value and magnitude are both equal length

    val <- numeric(length(value))

    for (i in 1:length(value)) {
        mag <- toupper(magnitude[i])
        
        if (grepl('[1-9]', mag)) {
            val[i] <- value[i] * 10^(as.numeric(mag) - 3)
        } else if (!grepl('[MB]', mag) | is.na(mag)) {
            val[i] <- value[i]
        } else if (mag == 'M') {
            val[i] <- value[i] * 1e3
        } else if (mag == 'B') {
            val[i] <- value[i] * 1e6
        }
    }

    val
}

dts[PROPDMG > 0, PropDmg:=dmgAmt(PROPDMG, PROPDMGEXP)]

dts[CROPDMG > 0, CropDmg:=dmgAmt(CROPDMG, CROPDMGEXP)]

dts[is.na(PropDmg), PropDmg:=0]
dts[is.na(CropDmg), CropDmg:=0]
```

### Summary from adjusting the affected damage estimates

```{r adjust magnitude}
bad.propexp <- dts[, PROPDMG > 0 & grepl('[^KBMkmb]', PROPDMGEXP)]

bad.cropexp <- dts[, CROPDMG > 0 & grepl('[^KMBkmb]', CROPDMGEXP)]

propdmg <- dts[bad.propexp, .(PROPDMG=sum(PROPDMG),
                              PropDmg=sum(PropDmg))]
cropdmg <- dts[bad.cropexp, .(CROPDMG=sum(CROPDMG),
                              CropDmg=sum(CropDmg))]

dmg.summary <- cbind(t(propdmg), t(cropdmg))
colnames(dmg.summary) <- c('Properties', 'Crops')
rownames(dmg.summary) <- c('Original', 'Adjusted')

dmg.summary
```

The adjustment resulted in `r ceiling(dmg.summary[2,1] / dmg.summary[1,1]) - 1`-fold
improvement in property damage estimates, or $`r ceiling((dmg.summary[2,1] - dmg.summary[1,1]) / 1e3)`
billions.


### Summary of damage estimates by states

```{r damage tally}
dmg.summary <- dts[, .('Prop Damage'=sum(PropDmg, na.rm = T),
                       'Crop Damage'=sum(CropDmg, na.rm = T),
                       Injuries=sum(INJURIES, na.rm = T),
                       Fatalitites=sum(FATALITIES, na.rm = T)),
                   by = STATE]

zero.dmg <- which(rowSums(dmg.summary[, c(2:5), with=F]) == 0)

head(dmg.summary)
```

And those states have no reported human injuries, fatalities, property or crop
damages are: `r dmg.summary$STATE[zero.dmg]`

### 2005 New Year's Eve Napa Valley Flood
The initial data exploration revealed that property and crop damage estimates of
the flood in Napa 2006 (REFNUM 605943) were erroneous by several order of
magnitude.

```{r napa original estimates}
#REFNUM 605943 original estimates
dts[REFNUM == 605943, c('PROPDMG', 'PROPDMGEXP', 'CROPDMG', 'CROPDMGEXP'), with=F]
```

 These values were corrected using estimated located in this
[News Article](http://napavalleyregister.com/news/local/napa-flood-damage-estimate-drops-to-million/article_bc2b3e27-1a7e-5886-ada4-6da209b4a5db.html 'Title'). 
[News footage](https://youtu.be/OQ5S86IAsn4) and additional information from
usgs.gov can be found [here](http://pubs.usgs.gov/of/2006/1182/pdf/ofr2006-1182.pdf).

```{r napa, results='hide'}
#REFNUM 605943 correction
dts[REFNUM == 605943, ':='(PROPDMG=13.75,
                           PROPDMGEXP='M',
                           CROPDMG=(114.2 - 13.75),
                           CROPDMGEXP='M')]
```



## Dimensionality reduction (feature extraction) of EVTYPE

```{r event types}
# Permitted Storm Data Events
evlist <- read.csv('EventTable.csv', na.strings='')

# Event types from the data set
evtype <- sort(dts[, unique(EVTYPE)])

head(evtype, 30)

```

Many of the EVTYPE (event types; `r length(evtype)` total) in the data set do
not match the [official list](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf)
of `r dim(evlist)[1]` event names. Best attempt was applied to semi-manually
pattern-match EVTYPE to the official list of events.

```
## EVTYPE text pattern 
#VOLCANIC ASH
dts[grepl('^vo', EVTYPE), EVTYPE:='Volcanic Ash']

#HAIL
dts[grepl('^hail', EVTYPE), EVTYPE:='Hail']

#HIGH WIND
dts[grepl('^high wind', EVTYPE), EVTYPE:='High Wind']

#HURRICANE
dts[grepl('^hurricane', EVTYPE), EVTYPE:='Hurricane (Typhoon)']
```
... however matching EVTYPE and maintaining the code this way gets tedious quickly.
A pragmatic alternative is to alphabetically sort and save the list
unique EVTYPE to a csv file first.

```{r write foo.csv}
write.table(sort(unique(dts$EVTYPE)),
            'foobar.csv', sep =',', col.names = c('EVTYPE'), row.names = F)
```

Then the official names of events are assigned to the adjacent column to produce
the a EVTYPE-to-EventName mapping file found [here](https://github.com/lawrence009/datasciencecoursera/blob/master/repdata/peer2/EventMapping.csv).

```{r merge evmap}
evmap <- read.csv('EventMapping.csv', na.strings = '', colClasses = 'character')[2:3]

#fraction of EVTYPE that is not mapped
sum(is.na(evmap)) / length(evtype)

#append the official event name
dts <- merge(dts, evmap, by = 'EVTYPE', all.x = T)

head(dts[, c('EVTYPE', 'EventName'), with=F])
```

Where ambiguous, EVTYPE is cross-reference against REMARKS.

```{r grepl EVTYPE REMARK, results='hide'}
## EVTYPE is summary...

#HUNDERSTORM WIND
dts[grepl('^summary', EVTYPE) & grepl('[Tt]hunderstorm.*[Ww]ind', REMARKS), EventName:='Thunderstorm Wind']

#HAIL
dts[grepl('^summary', EVTYPE) & grepl('[Hh]ail', REMARKS), EventName:='Hail']

#FLASH FLOOD
dts[grepl('^summary', EVTYPE) & grepl('[Ff]lash flood', REMARKS), EventName:='Flood']

#LIGHTNING
dts[grepl('^summary', EVTYPE) & grepl('[Ll]ightning', REMARKS), EventName:='Lightning']

#BLIZZARD
dts[grepl('^summary', EVTYPE) & grepl('[Bb]lizzard', REMARKS), EventName:='Blizzard']

```



### Assessment of EVTYPE extraction (map and replace) to EventName

```{r mapping assessment}

sum(is.na(dts$EventName)) / length(dts$EventName)

dmg.notmapped <- dts[is.na(EventName),
                    .(fatalities=sum(FATALITIES),
                      injuries=sum(INJURIES),
                      proddmg=sum(PropDmg),
                      cropdmg=sum(CropDmg))]


dmg.mapped <- dts[!is.na(EventName),
                  .(fatalities=sum(FATALITIES),
                    injuries=sum(INJURIES),
                    proddmg=sum(PropDmg),
                    cropdmg=sum(CropDmg))]


dmg.summary <- as.matrix(rbind(dmg.mapped, dmg.notmapped))
rownames(dmg.summary) <- c('mapped', 'not mapped')

dmg.summary

dmg.notmapped / dmg.mapped
```

**Conclusion:** The process is effective even though only `r round(sum(is.na(evmap)) / length(evtype) * 100)`
percent of the EVTYPE is mapped.

### Key weather elements

Weather events can also be further reduced using `r length(unique(evlist$Element1))`
key weather elements to enhance the analysis: `r sort(unique(evlist$Element1))`.

```{r elements}
summary(evlist) # list of official event names and the associated key elements

dts <- merge(dts, evlist, by = 'EventName', all.x = T)

summary(dts[, c('EVTYPE', 'EventName', 'Element1', 'Element2'), with=F])

```

The majority of the weather events are linked to key weather elements: water, ice and wind. 

### Bureau of Economic Analysis (BEA)
When comparing economic data, BEA further defines the states in the US into
[8 regions](http://bea.gov/regional/docs/regions.cfm):

1. New England
2. Mideast
3. Great Lakes
4. Plains
5. Southeast
6. Southwest
7. Rocky Mountain
8. Far West

The region-to-state mapping extracted from BEA can be found [here](https://github.com/lawrence009/datasciencecoursera/blob/master/repdata/peer2/BureauOfEconomicAnalysisRegions.csv).

```{r BEA US regions}
regions <- read.csv('BureauOfEconomicAnalysisRegions.csv')[c(3, 1)]
colnames(regions)[1] <- 'STATE'

summary(regions)

dts <- merge(dts, regions, by = 'STATE', all.x = T)

summary(dts[, c('STATE', 'Region'), with=F])
```

# Results

Weather events reported between 1950 and 2011 from the NOAA Storm Database were analyzed for the frequencies of the types of event and the portion of the those
event actually resulted in either human fatalities, injuries, property damages,
or corp damages.  The frequencies were further subdivided into regions and sorted
by event frequencies to aid prioritizing emergency resource and preparation.

Assessments of human causalities and economy impact by region and primary
weather elements were presented next.  A causality multiplier (lamda = 100) was
applied to the fatality figures (fatality = 100 injuries) when assessing impact
too population to health.

## Event frequencies

States in the Southeast regions are most likely to be susceptible to weather
event linked to wind, water or ice.  States in the Plains were most susceptible
to events linked to ice.

```{r fig1 event freq}
colnames(dts)

dts <- dts[, c(1:2, 5, 24:29, 38:43), with=F]

colnames(dts)

library(reshape2)
library(lattice)

dt1 <- table(dts[, c('Region', 'Element1'), with=F])
dt1 <- melt(dt1)

poly.colors <- trellis.par.get("superpose.polygon")$col
poly.colors <- c(poly.colors, 'plum')

barchart(value ~ Element1, data = dt1,
         main = 'Frequency of Events by Element',
         groups = Region,
         beside = T,
         ylab = 'No. of Occurences',
         scales=list(x=list(rot = 90)),
         auto.key=list(x = .05, y=.6, corner = c(0, 0), cex = 0.7),
         par.settings = list(superpose.polygon = list(col = poly.colors)))
```

## Frequencies of events with human fatalities by region

```{r}
library(knitr)

table(dts[(PropDmg + CropDmg) > 0, c('Region', 'Element1'), with=F]) -> dmg.freq
table(dts[FATALITIES > 0, c('Region', 'Element1'), with=F]) -> ftl.freq
table(dts[INJURIES   > 0, c('Region', 'Element1'), with=F]) -> inj.freq


freq.summary <- function(x) {
    # returns a data frame summarizing the frequency of element by region
    # x is a subset of dts that resembles y
    
    # distribution of event type by region
    y <- table(dts[, c('Region', 'Element1'), with=F])

    z <- merge(x, y, by=c('Region', 'Element1'))

    z$ratio <- z$Freq.x / z$Freq.y

    with(z, z[order(Region, -Freq.x, -ratio), ])
}


# percent of event with human fatalities
kable(freq.summary(ftl.freq), row.names = F, digits=2)
```

## Frequencies of events with human injuries by region

```{r}
# percent of event with human injuries
kable(freq.summary(inj.freq), row.names = F, digits=2)
```

## Frequencies of events with property or crop damages by region

```{r}
# percent of event with property or crop damage
kable(freq.summary(dmg.freq), row.names = F, digits=2)
 
```

## Impact to population health by weather elements and regions

Impact from heat was highest among all regions, followed by cold.

```{r fig2 humn}
library(lattice)

#casuality mulitplier
lamda <- 100

dts[, ':='(econ  =(PropDmg + CropDmg),
           humn  =(FATALITIES*lamda + INJURIES),
           year  =as.factor(year(BGN_DATE)),
           month =as.factor(month(BGN_DATE)))] -> dts #supress print; not copied

bwplot(humn~Element1 | Region , data = dts[humn > 0],
       par.strip.text = list(cex = 0.8),
       main = 'Human Casulities (adjusted for fatalities)',
       ylab = 'Fatalities and Injuries',
       scales = list(cex = 0.8,
                     x = list(rot = 90),
                     y = list(log = 10)))


```

## Impact to the economy by weather elements and regions

Except in Far West and New England states, heat on the average caused the most
economic damages.

```{r fig3 econ}

bwplot(econ~Element1 | Region, data = dts[econ > 0],
       par.strip.text = list(cex = 0.8),
       main = 'Economic Impact',
       ylab = 'Property and Crop Damages (dollar)',
       scales = list(cex = 0.8,
                     x = list(rot = 90),
                     y = list(log = 10)))


```

