---
title: "Manual Transmission Cars Deliver Better Fuel Economy"
author: "Lawrence Chen"
date: "January 31, 2016"
output: pdf_document
---

# Synopsis

You work for Motor Trend, a magazine about the automobile industry. Looking
at a data set of a collection of cars, they are interested in exploring the
relationship between a set of variables and miles per gallon (MPG) (outcome).
They are particularly interested in the following two questions:

* Is an automatic or manual transmission better for MPG?
* Quantify the MPG difference between automatic and manual transmissions

# Data set

The data was extracted from the 1974 Motor Trend US magazine, and comprises
fuel consumption and 10 aspects of automobile design and performance for 32
automobiles (1973–74 models). (see R {datasets} documentation)

```{r data}
# explicitly load mtcars from {datasets}
data(mtcars)
str(mtcars)

mtcars[, c('cyl', 'gear', 'carb')] <- lapply(mtcars[, c('cyl', 'gear', 'carb')], factor)
mtcars$vs <- factor(mtcars$vs, labels = c('V Engine', 'Straight Engine'))
mtcars$am <- factor(mtcars$am, labels = c('automatic', 'manual'))
```

# Exploratory data analyses

The first figure in Appendix suggests that manual transmission provides better
gas mileage.

The next figure suggests factors such drat, hp, cyl, disp, wt, vs and am have either
positive or negative correlation with mpg as shown in the next figure.

We will build a multivariable linear model to quantify the effects of these
variable on gas mileage.

# Model selection and regression analysis

'Backward' stepwise model selection to determine the variable for the best model.

*Note:*  Alternatively, a lengthier bidirectional stepwise selection and AIC
algorithm would arrive at the same model.

```{r model, results='hide'}
full.model <- lm(mpg ~ ., data = mtcars)
best.model <- step(full.model, direction = 'backward')
```

```{r}
summary(best.model)
```

The best model includes the cyl6, cyl8, hp, wt and ammanual variables with
the overall p-value < .001 with 84% of the variance explained by the final
model.

The model suggests that mpg decreases with respect to cylinders, horsepower,
and weight. As the earlier figure suggests mpg increases with respect to
having a manual transmission.

```{r}
t.test(mpg ~ am, mtcars)
```

p <= .001, and therefore accepting the alternative hypothesis that manual
transmission offer better gas mileage.

# Conclusions
* The cars with a manual transmission are better for mpg than cars with an automatic transmission
* The rate of change of the conditional mean mpg with respect to am is about 1.8,
with 95% confident that this value varies between -1.06 and 4.68. 
* If the number of cylinders increase from 4 to 8, mpg will decrease by a factor of 3

# Appendix
```{r}
boxplot(mpg ~ am, data = mtcars,
        xlab = 'Transmission', ylab = 'MPG',
        main = paste0('Fuel Economy of Automatic vs. Manual Transmission\n(n=',
                      nrow(mtcars), ')'))
pairs(mtcars)

plot(best.model)

plot(best.model, which=c(4,6))
```


---

*Note:* See https://github.com/lawrence009/jhudatascience/tree/master/reg for the source Rmd.

