---
title: 'Statistical Inference Course Project: Part One (Simulation Exercise)'
author: "Lawrence Chen"
date: "January 31, 2016"
output: pdf_document
---

# Overview

The following exercise investigates the properties of the exponential
distribution via simulation and compares it with the Central Limit Theorem.
The simulation will show that the the averages of 40 exponentials is 
approximately normal.

The exponential distribution is expressed in R as `rexp(n, lamda)` where
`lambda` is the rate parameter. The mean of exponential distribution is
`1/lambda` and the standard deviation is also `1/lambda`.

lambda is set at 0.2 for the sample (n=40) given 1,000 simulated averages.

```{r sim}
set.seed(25)

# simulation vars
lambda <- .2
n <- 40L
sims <- 1e3L

# averages of 40 exponetials given 1,000 simluations
sim.means <- rowMeans(matrix(rexp(sims * n, rate = lambda), sims, n))

summary(sim.means)
```

# Results

## Center of distribution

The sample mean of 40 exponentials is centered at `r round(mean(sim.means), 3)`,
which is very close to `r round(1/lambda, 3)`, the theoretical center of
distribution 1/lambda.

## Sample variability

```{r vars}
# sample sd
sd(sim.means)

# expected sd
(1/lambda) / sqrt(n)

# variance of the sample
var(sim.means)

# variance of the expected
((1/lambda) / sqrt(n))^2
```
The variabilities in the distribution of the sample are very close to the
theoretical variance, i.e., Central Limit Theorem.  

## Approximately normal sample distribution

The sample distribution (graphically illustrated as white bars) are
approximately normally distributed, as compared to theoretical normal
(demonstrated via `dnorm`).

```{r hist}
library(ggplot2)

ggplot(data.frame(x=sim.means), aes(x=sim.means)) +
  geom_histogram(aes(y=..density..),
                 fill = 'white', color = 'black', 
                 binwidth = .2) +
  stat_function(fun = dnorm,
                arg = list(mean = 1/lambda, sd = (1/lambda) / sqrt(n))) +
  labs(x = 'Average of 40 Samples',
       title = 'Sample vs. Theroritical Disitrubtions')
```


---

*Note:* See https://github.com/lawrence009/jhudatascience/tree/master/statinf for the source Rmd.

