---
title: 'Statistical Inference Course Project: Part Two (the Response of Vitamin C on
  Tooth Growth in Guinea Pigs)'
author: "Lawrence Chen"
date: "January 31, 2016"
output: pdf_document
---


# Experiment results

Sixty guinea pigs were treated with vitamin C which suggests its positive response on
tooth growth. The data is sourced from C. I. Bliss (1952)
*The Statistics of Bioassay*. Academic Press.

The response is measured as the length of odontoblasts (cells responsible for
tooth growth). Each animal received one of three dose levels of vitamin C
(0.5, 1, and 2 mg/day) by one of two delivery methods. The control group
receives orange juice (coded as OJ); the experimental group received ascorbic
acid (a form of vitamin C and coded as VC).

```{r data, echo=FALSE}
# Explicitly load ToothGrowth data.frame from {datasets}
data(ToothGrowth)

library(ggplot2)
ggplot(ToothGrowth, aes(x = as.factor(dose), y = len, color = supp)) +
  geom_boxplot() +
  labs(x = 'Dose (mg/Day)', y = 'Response')
```

The data seems to suggest that OJ had a more positive effect on tooth growth
at lower dosage and that the effect diminished as dosage increased.

```{r}
summary(ToothGrowth)
with(ToothGrowth, tapply(len, list(supp, dose), mean))
with(ToothGrowth, tapply(len, list(supp, dose), var))
```

# Hyopthesis testing

## H~0~: Dosage, independent of the types of supplements, has no signficant effect.

```{r}
with(subset(ToothGrowth, dose == c(0.5, 1)), t.test(len~dose))
```

p < 0.01 and the confidence intervals clearly do not overlap the group means.
We therefore conclude there is a positive effect of supplement dosage.

*Omitted:* additional t-tests can be performed against other dosages to
further validate this point.

## H~0~: There is no difference in the effect between two suppliments independent of dosage. 

```{r}
with(ToothGrowth, t.test(len~supp))
```

p > 0.05. Therefore we accept the null hypothesis that both supplements
have equal effect on tooth growth

## H~0~: There is no difference in the effect at lower dosages.

```{r}
with(subset(ToothGrowth, dose == 1), t.test(len~supp))
with(subset(ToothGrowth, dose == .5), t.test(len~supp))
```

p < 0.05. However, the confidence intervals overlap the two group means, and
therefore cannot definitively conclude there is significant difference
at lower dosages.


# Conclusions

* Higher dosage, regardless of the source of vitamin C, suggests providing
supplements promotes tooth growth
* There is no significant difference between using OJ or VC as supplements
* In order to determine if vitamin C promotes tooth growth, the experiment
should be repeated with a control group given a placebo, i.e. no vitamin C 

---

*Note:* See https://github.com/lawrence009/jhudatascience/tree/master/statinf for the source Rmd.
